Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-03-06 paper_comparison/paper_comparison/paper_comparison.tex
update the effect size desc

@@

Graphical summaries in which residuals are plotted against fitted values
or other functions of the regressors that are approximately orthogonal
to residuals are referred to as standard residual plots in
\citet{cook1982residuals}. The first panel of Figure
\ref{fig:residual-plot-common-departures} shows an ideal residual plot
with residuals evenly distributed at both sides of the horizontal zero
line, with no noticeable patterns.

@@

Non-linearity is a type of model misspecification caused by failing to
include higher order terms of the regressors in the regression equation.
Any non-linear functional form of residuals on fitted values in the
residual plot could be indicative of non-linearity. An example residual
plot containing visual pattern of non-linearity is given at the second
panel of Figure \ref{fig:residual-plot-common-departures}. One can
clearly observe the ``S-shape'' from the residual plot as the cubic term
is not captured by the misspecified model.

@@

Other than checking diagnostic plots, analysts may perform formal
hypothesis testing for detecting model defects. Depending on the
alternative hypothesis that is focused on, a variety of tests can be
applied. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the Breusch-Pagan
test \citep{breusch_simple_1979}, which are both derived from the
Lagrange multiplier test \citep{silvey1959lagrangian} principle that
relies on the asymptotic properties of the null distribution. For
testing non-linearity, one may apply the F-test as a model structural
test to examine the significance of specific polynomial and non-linear
forms of the regressors, or the significance of proxy variables as in
the Ramsey Regression Equation Specification Error Test (RESET)
\citep{ramsey_tests_1969}. The Shapiro-Wilk test
\citep{shapiro1965analysis} is the most widely used test of
non-normality included by many of the statistical softwares. The
Jarque--Bera test \citep{jarque1980efficient} is also used to directly
checks if the sample skewness and kurtosis match a normal distribution.

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding RESET test, Breusch-Pagan test and Shapiro-Wilk test as
shown in Table \ref{tab:example-residual-plot-table}. In the example,
the Breusch-Pagan test and the Shapiro--Wilk test both reject the null
hypothesis \(H_0\) for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are also sensitive
to other types of departures. It is likely \(H_0\) is correctly rejected
but for the wrong reason, a phenomenon known as the ``Type III error''.
Additionally, outliers will often incorrectly trigger the rejection of
\(H_0\) despite when majority of the residuals are well-behaved
\citep{cook_applied_1999}. Furthermore, with a sufficiently large sample
size, residual-based tests may reject \(H_0\) due to a slight departure
that is of little practical significance. These can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers
and slight departures.

@@

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) null plots have the identical graphical procedure except the
data has been replaced with data consistent with \(H_0\). Then, an
observer who have not seen the data plot will be asked to point out the
most different plot from the lineup. Under \(H_0\), it is expected that
the data plot would have no distinguishable difference from the null
plots, and the probability that the observer correctly picks the data
plot is \(1/m\). If one rejects \(H_0\) as the observer correctly picks
the data plot, then the Type I error of this test is \(1/m\).

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the data plot at position \(2^2 + 2\) is identifiable, then
it is evidence for the rejection of \(H_0\) that the regression model is
correctly specified. In fact, the actual residual plot is obtained from
a misspecified regression model with non-linearity defect.

Data used in the \(m - 1\) null plots needs to be simulated. In
regression diagnostics, sampling data consistent with \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
regressors, then rescaling it by the ratio of residual sum of square in
two regressions.

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two regressors. Their results suggest that visual
tests are capable of testing the significance of a single regressor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. (contradict to our claim? expand?) In their third
experiment, where there is not a conventional test, visual test
outperforms the conventional test for a large margin. This is
encouraging, as it promotes the use of visual inference in situations
where there are no existing statistical testing procedures. Visual
inference have also been integrated into diagnostic of hierarchical
linear models by \citet{loy2013diagnostic}, \citet{loy2014hlmdiag} and
\citet{loy2015you}. They use lineup protocols to judge the assumption of
linearity, normality and constant error variance for both the level-1
and level-2 residuals. (expand?)

@@

However, it is practically impossible to test this specific \(H_0\) with
conventional tests, which are constructed to measure specific
departures. For example, the RESET test is formulated as
\(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\),
from
\(e = \gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\).
Similarly, the BP test is designed to specifically test \(H_0:\)
\emph{error variances are all equal} (\(\tau=0\)) versus the alternative
\(H_a:\) \emph{that the error variances are a multiplicative function of
one or more variables} (\(\tau\neq 0\)) from
\(e^2 = \tau\hat{y} + u, ~ u\sim N(0,\sigma_u^2)\).

@@

The power of a model misspecification test is the probability that the
null hypothesis is rejected given the regression model is misspecified
in a specific way. It is an important indicator when one is concerned
about whether model assumptions have been violated. Although in
practice, one might be more interested in knowing how much the residuals
deviate from the model assumptions, and whether this deviation is of
practical significance.

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
effect size is provided in Section \ref{effect-size} after the
introduction of the simulation model used in our human subject
experiment. The theoretical power of a test is sometimes not a trival
solution. For a given regression model under a specific effect size, if
the data generating process is also known, the power of a conventional
test can be estimated by counting the number of times the test rejects
the null hypothesis in a repeated sampling exercise and divided by the
number of times simulated from the known data generating process. This
is the method we used to estimate the power of conventional tests in our
experiment.

The power of a visual test on the other hand, may additionally depend on
the ability of the particular subject, as the skill of the individual
may affect the number of observers who identify the data plot from the
lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of a subject
\(j\) correctly picking the data plot from a lineup \(l\) using a
mixed-effect logistic regression, with subjects treated as random
effect. Then, the estimated power of a visual test evaluated by a single
subject is the predicted value obtained from the mix-effect model.

However, this mix-effect model does not work with scenario where
subjects are asked to select one or more ``most different'' plots. In
this scenario, having the probability of a subject \(j\) correctly
picking the data plot from a lineup \(l\) is insufficient to determine
the power of a visual test because it does not provide information about
the number of selections made by the subject for the calculation of the
p-value (See Equation \ref{eq:pvalue-beta-binomial-approx}).

Therefore, we directly estimate the probability of a lineup being
rejected by assuming that individual skill has negligible effect on the
variation of the power. This assumption is not necessary true, but it
helps simplifying the model structure, thereby obviate a costly
large-scale experiment to estimate complex covariance matrices. There is
a discussion provided in the Appendix about a potential solution to
model individual effect in the multiple selections scenario.

(should we just link the power together for conventional test?)

We use the same logistic regression model to fit the power of
conventional tests and visual tests formulated as

@@

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = exp(z)/(1+exp(z))\). The effect size \(E\) is the only
regressor and the intercept is fixed to \(log(0.05/0.95)\) so that
\(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\).

To study various factors contributing to the power of test, the same
logistic regression model is fit on different subsets of the collated
data grouped by levels of factors. These include the distribution of the
regressor, type of the simulation model and the shape of the residual
departures.

@@

To investigate the difference between conventional hypothesis testing
and visual inference in the application of linear regression
diagnostics, an experiment is conducted in three data collection
periods. Two types of departures, non-linearity and heteroskedasticity,
are primarily collected during data collection periods I and II. The
data collection period III was designed primarily to measure human
responses to null lineups so that the parameter \(\alpha\) in Equation
\ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for
both non-linearity and heteroskedasticity, using uniform \(X\)
distribution, were included so that the participants were evaluating
some lineups with signal also. It would be too frustrating for
participants to only be assigned lineups with all null plots. Overall,
we collected 7974 evaluations on 1152 unique lineups performed by 443
subjects throughout three data collection periods.

@@

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups
used in non-linearity detection. This lineup is produced by the
non-linearity model with \(j = 6\) and uniform fitted values. The data
plot location is \(2^3 - 4\). All five subjects correctly identify the
data plot from this lineup.

@@

\caption{Examining the effect of signal strength for the three different values of $n$ used in the experiment, for non-linear structure with fixed $\sigma = 1.5$, uniform $X$ distribution, and S shape. For these factor levels, only when $n=300$ is the S shape clearly visible.}\label{fig:different-n}
\end{figure}

@@

\caption{Examining the effect of $\sigma$ on the signal strength in the non-linearity detection, for $n=300$, uniform $X$ distribution and the U shape. As $\sigma$ increases the signal strength decreases, to the point that the U is almost unrecognisable when $\sigma=4$.}\label{fig:different-sigma}
\end{figure}

@@

For \(b \neq 0\), the variance-covariance matrix of the error term
\(\boldsymbol{\varepsilon}\) is correlated with the regressor
\(\boldsymbol{x}\), which will lead to the presence of
heteroskedasticity. Visual patterns of heteroskedasticity are simulated
using three different shapes (\(a\) = -1, 0, 1). (A summary of the
factors can be found in Table \ref{tab:model-factor-table}.)

@@

An example lineup of this model used in data collection period II is
shown in Figure \ref{fig:example-heter-lineup} with \(a = -1\) and
\(X_{raw} \sim U(-1, 1)\). The data plot location is \(2^4 + 2\). Nine
out of 11 subjects correctly identify the data plot from this lineup.

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. The
distribution may be relatively Gaussian, reflecting a linear combinatio
of many predictors, adhering to the Central Limit Theorem. It is also
common to see a skewed distribution of fitted values, if one or more of
the predictors has a skewed distribution. This latter problem is usually
corrected before modeling using a variable transformation. Our
simulation asesses this by using four different distributions to
represent the fitted values (\(X_{raw}\)): (1) \(U(-1, 1)\), (2)
\(N(0, 0.3^2)\), (3) \(lognormal(0, 0.6^2)/3\) and (4) \(u\{1, 5\}\). We
would expect that the best reading of residual plots occurs when the
fitted values are uniformly distributed.

Three different sample sizes are used, \(n=50, 100, 300\) across the
experiments. We would expect considerable variation in the signal
strength in the simulated data plots with smaller \(n\). A sample size
of 300 is typically enough for structure to be visible in a scatterplot
reliably.

@@

As summarised in Table \ref{tab:model-factor-table}, three additional
parameters \(n\), \(\sigma\) and \(b\) are used to control the strength
of the signal so that different difficulty levels of lineups are
generated, and therefore, the estimated power curve will be smooth and
continuous. Parameter \(\sigma \in \{0.5, 1, 2, 4\}\) and
\(b \in \{0.25, 1, 4, 16, 64\}\) are used in data collection periods I
and II respectively. Figure \ref{fig:different-sigma} and
\ref{fig:different-b} demonstrate the impact of these two parameters. A
large value of \(\sigma\) will increase the variation of the error of
the non-linearity model and decrease the visibility of the visual
pattern. The parameter \(b\) controls the standard deviation of the
error across the support of the regressor. Given \(x \neq a\), a larger
value of \(b\) will lead to a larger ratio of the variance at \(x\) to
the variance at \(x - a = 0\), making the visual pattern more obvious.

@@

(do we actually need the penalty term?)

Effect size \(E\) can be defined as the difference of a parameter for a
particular model or distribution, or a statistic derived from a sample.
Importantly, it needs to reflect the treatment we try to measure.
According to the underlying simulated model, we define the effect size
\(E = \sqrt{n}/(\sigma k)\) for the non-linearity model and
\(E = b\sqrt{n}/(|a| + 1)\) for the heteroskedasticity model. The
parameter \(k\) is the number of turning points embedded in the residual
plot. It is used as a penalty term to reduce the effect size for complex
shapes. There are one, two, three and five turning points in ``U'',
``S'', ``M'', ``Triple-U'' shape respectively. The denominator of the
effect size of the heteroskedasticity model is \(|a| + 1\). It is used
to

@@

As mentioned in Section \ref{calculating-p-values-for-the-visual-test},
\(\alpha\) used in Equation \ref{eq:pvalue-beta-binomial} needs to be
estimated using null lineups. Hence, three replications of 36 lineups
with all combinations of \(n\) and fitted value distributions are
included in data collection period III. In these lineups, the data of
the data plot is generated from a model with zero effect size, while the
data of the 19 null plots are generated using the same simulation method
discussed in Section \ref{sampling-from-the-null-distribution}. This
generation procedure differs from the canonical Rorschach lineup
procedure, which requires that all 20 plots are generated from the null
hypothesis. However, these lineups serve the same fundamental purpose:
to assess the number of visually interesting plots generated under the
null hypothesis.

@@

We pick RESET test (\texttt{resettest}) and Breusch-Pagan test
(\texttt{bptest}) from the R package \texttt{lmtest}, and Shapiro-Wilk
test (\texttt{shapiro.test}) from the built-in R package \texttt{stats}.
Among them, RESET test is the only exact and appropriate test in this
scenario. Both the Breusch-Pagan test and the Shapiro-Wilk test are
approximate and inappropriate tests. Their estimated power is shown in
Figure \ref{fig:polypower}. To set up the RESET test, we include
different powers of fitted values as proxies. According to
\citet{ramsey_tests_1969}, there are no general rules for the power of
the fitted values needed by the RESET test, but it finds power up to
four is usually sufficient. Thus, we follow this guideline to conduct
the RESET test. For the Breusch-Pagan test, the choice of regressors in
the auxiliary regression is left to the user
\citep{breusch_simple_1979}. But as \citet{waldman1983note} suggested,
it is a good choice for the set of auxiliary regressors in the
Breusch-Pagan test be the same as the White test. Thus, we include both
\(\boldsymbol{x}\) and \(\boldsymbol{x}^2\) in the auxiliary regression.

@@

For both the non-linearity pattern and the heteroskedasticity pattern,
the power curves of appropriate conventional tests are constantly
greater than the power curves of visual tests and other inappropriate
conventional tests. At \(log(E) = 1\) for non-linearity model, where the
residual departures are unnoticeable from the corresponding residual
plot, the RESET test has above 25\% of chance of rejecting the null
hypothesis, while only a few visual tests would result in a rejection.
Similarly, at \(log(E) = 3\) for heteroskedasticity model, the BP test
has around 25\% power but the non-constant variance visual pattern is
almost unobserved from the residual plot and the power of the visual
test is close to the significance level at this point.

>>>>

(title refine - rank: low)

(fix the tense issue - rank: low)

(fix the notation issue (bold?) - rank: low)

@@

Graphical summaries in which residuals are plotted against fitted values
or other functions of the predictors that are approximately orthogonal
to residuals are referred to as standard residual plots in
\citet{cook1982residuals}. The first panel of Figure
\ref{fig:residual-plot-common-departures} shows an ideal residual plot
with residuals evenly distributed at both sides of the horizontal zero
line, with no noticeable patterns.

@@

Non-linearity is a type of model misspecification caused by failing to
include higher order terms of the predictors in the regression equation.
Any non-linear functional form of residuals on fitted values in the
residual plot could be indicative of non-linearity. An example residual
plot containing visual pattern of non-linearity is given at the second
panel of Figure \ref{fig:residual-plot-common-departures}. One can
clearly observe the ``S-shape'' from the residual plot as the cubic term
is not captured by the misspecified model.

@@

Other than checking diagnostic plots, analysts may perform formal
hypothesis testing for detecting model defects. Depending on the
alternative hypothesis that is focused on, a variety of tests can be
applied. For example, the presence of heteroskedasticity can usually be
tested by applying the White test
\citep{white_heteroskedasticity-consistent_1980} or the Breusch-Pagan
test \citep{breusch_simple_1979}, which are both derived from the
Lagrange multiplier test \citep{silvey1959lagrangian} principle that
relies on the asymptotic properties of the null distribution. For
testing non-linearity, one may apply the F-test as a model structural
test to examine the significance of specific polynomial and non-linear
forms of the predictors, or the significance of proxy variables as in
the Ramsey Regression Equation Specification Error Test (RESET)
\citep{ramsey_tests_1969}. The Shapiro-Wilk test
\citep{shapiro1965analysis} is the most widely used test of
non-normality included by many of the statistical softwares. The
Jarque--Bera test \citep{jarque1980efficient} is also used to directly
checks if the sample skewness and kurtosis match a normal distribution.

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding RESET test, Breusch-Pagan test and Shapiro-Wilk test as
shown in Table \ref{tab:example-residual-plot-table}. In the example,
the Breusch-Pagan test and the Shapiro--Wilk test both reject \(H_0\)
for departures that they do not intend to examine. As discussed in
\citet{cook1982residuals}, most residual-based tests for a particular
type of departure from model assumptions are also sensitive to other
types of departures. It is likely \(H_0\) is correctly rejected but for
the wrong reason, a phenomenon known as the ``Type III error''.
Additionally, outliers will often incorrectly trigger the rejection of
\(H_0\) despite when majority of the residuals are well-behaved
\citep{cook_applied_1999}. Furthermore, with a sufficiently large sample
size, residual-based tests may reject \(H_0\) due to a slight departure
that is of little practical significance. These can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers
and slight departures.

@@

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots. Later,
\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test inspired by the ``police lineup'' or ``identity parade'' which is
the act of asking the eyewitness to identify criminal suspect from a
group of irrelevant people. The protocol consists of \(m\) randomly
placed plots, where one plot is the data plot, and the remaining
\(m - 1\) null plots have the identical graphical procedure except the
data has been replaced with data consistent with \(H_0\), that the model
is correctly specified. Then, an observer who have not seen the data
plot will be asked to point out the most different plot from the lineup.
Under \(H_0\), it is expected that the data plot would have no
distinguishable difference from the null plots, and the probability that
the observer correctly picks the data plot is \(1/m\). If one rejects
\(H_0\) as the observer correctly picks the data plot, then the Type I
error of this test is \(1/m\).

@@

Figure \ref{fig:first-example-lineup} is an example of a lineup
protocol. If the data plot at position \(2^2 + 2\) is identifiable, then
it is evidence for the rejection of \(H_0\). In fact, the actual
residual plot is obtained from a misspecified regression model with
missing non-linear terms.

Data used in the \(m - 1\) null plots needs to be simulated. In
regression diagnostics, sampling data consistent with \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually a composite
hypothesis controlled by nuisance parameters. Since regression models
can have various forms, there is no general solution to this problem,
but it sometimes can be reduced to a so called ``reference
distribution'' by applying one of the three methods: (i) sampling from a
conditional distribution given a minimal sufficient statistic under
\(H_0\), (ii) parametric bootstrap sampling with nuisance parameters
estimated under \(H_0\), and (iii) Bayesian posterior predictive
sampling. The conditional distribution given a minimal sufficient
statistic is the best justified reference distribution among the three
\citep{buja_statistical_2009}. Essentially, null residuals can be
simulated by regressing \(N\) i.i.d standard normal random draws on the
predictors, then rescaling it by the ratio of residual sum of square in
two regressions.

The effectiveness of lineup protocol for regression analysis is
validated by \citet{majumder_validation_2013} under relatively simple
settings with up to two predictors. Their results suggest that visual
tests are capable of testing the significance of a single predictor with
a similar power as a t-test, though they express that in general it is
unnecessary to use visual inference if there exists a conventional test,
and they do not expect the visual test to perform equally well as the
conventional test. In their third experiment, where there is not a
conventional test, visual test outperforms the conventional test for a
large margin. This is encouraging, as it promotes the use of visual
inference in situations where there are no existing statistical testing
procedures. Visual inference have also been integrated into diagnostic
of hierarchical linear models by \citet{loy2013diagnostic},
\citet{loy2014hlmdiag} and \citet{loy2015you}. They use lineup protocols
to judge the assumption of linearity, normality and constant error
variance for both the level-1 and level-2 residuals. (expand?)

@@

However, it is practically impossible to test this specific \(H_0\) with
conventional tests, which are constructed to measure specific
departures. For example, the RESET test is formulated as
\(H_0:\gamma_1 = \gamma_2 = \gamma_3 = 0\) against
\(H_a: \gamma_1 \neq 0 \text{ or } \gamma_2 \neq 0 \text{ or } \gamma_3 \neq 0\),
from
\(e = \tau_0 + \tau_1x_1 + ... + \tau_1 x_p +\gamma_1\hat{y}^2 + \gamma_2\hat{y}^3 + \gamma_3\hat{y}^4 + u, ~~u \sim N(0, \sigma_u^2)\).
Similarly, the Breusch-Pagan test is designed to specifically test
\(H_0:\) \emph{error variances are all equal}
(\(\tau_i=0 \text{ for } i=1,..,p\)) versus the alternative \(H_a:\)
\emph{that the error variances are a multiplicative function of one or
more variables} (\(\text{at least one } \tau_i\neq 0\)) from
\(e^2 = \tau_0 + \tau_1 x_1 + \cdots + \tau_1 x_p + u, ~ u\sim N(0,\sigma_u^2)\).

@@

The power of a model misspecification test is the probability that
\(H_0\) is rejected given the regression model is misspecified in a
specific way. It is an important indicator when one is concerned about
whether model assumptions have been violated. Although in practice, one
might be more interested in knowing how much the residuals deviate from
the model assumptions, and whether this deviation is of practical
significance.

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
effect size is provided in Section \ref{effect-size} after the
introduction of the simulation model used in our human subject
experiment. The theoretical power of a test is sometimes not a trival
solution, but it can be estimated if the data generating process is
known. We use a predefined model to generate a large set of simulated
data under different effect sizes, and record if the conventional test
rejects \(H_0\). The probability of the conventional test rejects
\(H_0\) is then fitted by a logistic regression formulated as

@@

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = exp(z)/(1+exp(z))\). The effect size \(E\) is the only
predictor and the intercept is fixed to \(log(0.05/0.95)\) so that
\(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\), which is the desired
significance level.

The power of a visual test on the other hand, may additionally depend on
the ability of the particular subject, as the skill of the individual
may affect the number of observers who identify the data plot from the
lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of a subject
\(j\) correctly picking the data plot from a lineup \(l\) using a
mixed-effect logistic regression, with subjects treated as random
effect. Then, the estimated power of a visual test evaluated by a single
subject is the predicted value obtained from the mix-effect model.
However, this mix-effect model does not work with scenario where
subjects are asked to select one or more most different plots. In this
scenario, having the probability of a subject \(j\) correctly picking
the data plot from a lineup \(l\) is insufficient to determine the power
of a visual test because it does not provide information about the
number of selections made by the subject for the calculation of the
p-value (See Equation \ref{eq:pvalue-beta-binomial-approx}). Therefore,
we directly estimate the probability of a lineup being rejected by
assuming that individual skill has negligible effect on the variation of
the power. This assumption is not necessary true, but it helps
simplifying the model structure, thereby obviate a costly large-scale
experiment to estimate complex covariance matrices. The same model given
in Equation \ref{eq:logistic-regression-1-1} is applied to model the
power of a visual test.

To study various factors contributing to the power of both tests, the
same logistic regression model is fit on different subsets of the
collated data grouped by levels of factors. These include the
distribution of the fitted values, type of the simulation model and the
shape of the residual departures.

@@

An experiment is conducted in three data collection periods to
investigate the difference between conventional hypothesis testing and
visual inference in the application of linear regression diagnostics.
Two types of departures, non-linearity and heteroskedasticity, are
collected during data collection periods I and II. The data collection
period III was designed primarily to measure human responses to null
lineups so that the parameter \(\alpha\) in Equation
\ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for
both non-linearity and heteroskedasticity, using uniform fitted value
distribution, were included so that the participants were evaluating
some lineups with signal also. It would be too frustrating for
participants to only be assigned lineups with all null plots. Overall,
we collected 7974 evaluations on 1152 unique lineups performed by 443
subjects throughout three data collection periods.

@@

Figure \ref{fig:example-poly-lineup} demonstrates one of the lineups
used in non-linearity detection. This lineup is produced by the
non-linearity model with \(j = 6\). The data plot location is
\(2^3 - 4\). All five subjects correctly identify the data plot from
this lineup.

@@

\caption{Examining the effect of signal strength for the three different values of $n$ used in the experiment, for non-linear structure with fixed $\sigma = 1$, uniform fitted value distribution, and S shape. For these factor levels, only when $n = 300$ is the S shape clearly visible.}\label{fig:different-n}
\end{figure}

@@

\caption{Examining the effect of $\sigma$ on the signal strength in the non-linearity detection, for $n=300$, uniform fitted value distribution and the U shape. As $\sigma$ increases the signal strength decreases, to the point that the U is almost unrecognisable when $\sigma=4$.}\label{fig:different-sigma}
\end{figure}

@@

For \(b \neq 0\), the variance-covariance matrix of the error term
\(\boldsymbol{\varepsilon}\) is correlated with the predictor
\(\boldsymbol{x}\), which will lead to the presence of
heteroskedasticity. Visual patterns of heteroskedasticity are simulated
using three different shapes (\(a\) = -1, 0, 1). (A summary of the
factors can be found in Table \ref{tab:model-factor-table}.)

@@

An example lineup of this model used in data collection period II is
shown in Figure \ref{fig:example-heter-lineup} with \(a = -1\). The data
plot location is \(2^4 + 2\). Nine out of 11 subjects correctly identify
the data plot from this lineup.

@@

Fitted values are a function of the independent variables, and the
distribution of the observed values affects the distribution of the
fitted values. In the best case scenario the fitted values will have a
uniform distribution, which means that there is even coverage of
possible observed values across all of the predictors. This is not
always present in the collected data. Sometimes the fitted values are
discrete because one or more predictors were measured discretely. The
distribution may be relatively Gaussian, reflecting a linear combinatio
of many predictors, adhering to the Central Limit Theorem. It is also
common to see a skewed distribution of fitted values, if one or more of
the predictors has a skewed distribution. This latter problem is usually
corrected before modelling using a variable transformation. Our
simulation assess this by using four different distributions to
represent fitted values: (1) uniform, (2) normal, (3) skewed and (4)
discrete uniform. This is constructed by defining the raw predictor
\(X_{raw}\) in four corresponding distributions: (1) \(U(-1, 1)\), (2)
\(N(0, 0.3^2)\), (3) \(lognormal(0, 0.6^2)/3\) and (4) \(u\{1, 5\}\). We
would expect that the best reading of residual plots occurs when the
fitted values are uniformly distributed.

Three different sample sizes are used, \(n=50, 100, 300\) across the
experiments. We would expect considerable variation in the signal
strength in the simulated data plots with smaller \(n\). A sample size
of 300 is typically enough for structure to be visible in a scatter plot
reliably.

@@

As summarised in Table \ref{tab:model-factor-table}, three additional
parameters \(n\), \(\sigma\) and \(b\) are used to control the strength
of the signal so that different difficulty levels of lineups are
generated, and therefore, the estimated power curve will be smooth and
continuous. Parameter \(\sigma \in \{0.5, 1, 2, 4\}\) and
\(b \in \{0.25, 1, 4, 16, 64\}\) are used in data collection periods I
and II respectively. Figure \ref{fig:different-sigma} and
\ref{fig:different-b} demonstrate the impact of these two parameters. A
large value of \(\sigma\) will increase the variation of the error of
the non-linearity model and decrease the visibility of the visual
pattern. The parameter \(b\) controls the standard deviation of the
error across the support of the predictor. Given \(x \neq a\), a larger
value of \(b\) will lead to a larger ratio of the variance at \(x\) to
the variance at \(x - a = 0\), making the visual pattern more obvious.

@@

Effect size \(E\) can be defined as the difference of a parameter for a
particular model or distribution, or a statistic derived from a sample.
Importantly, it needs to reflect the treatment we try to measure.
According to the underlying simulated model, we define the effect size

\[E = \frac{1}{2}\left(\boldsymbol{\mu}_z'[diag(\boldsymbol{R}_a\sigma^2)]^{-1}\boldsymbol{\mu}_z\right)\]

for the non-linearity model, and

\[E = \frac{1}{2}\left(log\frac{|diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')|}{|diag(\boldsymbol{R}_a)|} - n + tr(diag(\boldsymbol{R}_a\boldsymbol{V}\boldsymbol{R}_a')^{-1}diag(\boldsymbol{R}_a))\right)\]
for the heteroskedasticity model, where \(diag(.)\) is the diagonal
matrix constructed from the diagonal elements of a matrix,
\(\sigma^2\boldsymbol{I}\) is the assumed covariance matrix of the error
term when \(H_0\) is true, \(\boldsymbol{V}\) is the actual covariance
matrix of the error term,
\(\boldsymbol{R}_a = \boldsymbol{I}_n - \boldsymbol{H}_a\) is the
residual operator,
\(\boldsymbol{H}_a = \boldsymbol{X}_a(\boldsymbol{X}_a'\boldsymbol{X}_a)^{-1}\boldsymbol{X}_a'\)
is the hat matrix,
\(\boldsymbol{X}_a = (\boldsymbol{1}, \boldsymbol{X})\) is the matrix of
predictors including the intercept used in the regression equation, and
\(\boldsymbol{\mu}_z = \boldsymbol{R}_a\boldsymbol{Z}\boldsymbol{\beta}_z\)
is the expected values of residuals with \(\boldsymbol{Z}\) be any
higher order terms of \(\boldsymbol{X}\) leave out by the regression
equation and \(\boldsymbol{\beta}_z\) be the corresponding coefficients.

Both equations above are derived from the Kullback-Leibler divergence
(citation here?) and the details are provided in the appendix.

@@

As mentioned in Section \ref{calculating-p-values-for-the-visual-test},
\(\alpha\) used in Equation \ref{eq:pvalue-beta-binomial} needs to be
estimated using null lineups. Hence, three replications of 36 lineups
with all combinations of \(n\) and fitted value distributions are
included in data collection period III. In these lineups, the data of
the data plot is generated from a model with zero effect size, while the
data of the 19 null plots are generated using the same simulation method
discussed in Section \ref{sampling-from-the-null-distribution}. This
generation procedure differs from the canonical Rorschach lineup
procedure, which requires that all 20 plots are generated from \(H_0\).
However, these lineups serve the same fundamental purpose: to assess the
number of visually interesting plots generated under \(H_0\).

@@

We pick RESET test (\texttt{resettest}) and Breusch-Pagan test
(\texttt{bptest}) from the R package \texttt{lmtest}, and Shapiro-Wilk
test (\texttt{shapiro.test}) from the built-in R package \texttt{stats}.
Among them, RESET test is the only exact and appropriate test in this
scenario. Both the Breusch-Pagan test and the Shapiro-Wilk test are
approximate and inappropriate tests. Their estimated power is shown in
Figure \ref{fig:polypower}. To set up the RESET test, we include
different powers of fitted values as proxies. According to
\citet{ramsey_tests_1969}, there are no general rules for the power of
the fitted values needed by the RESET test, but it finds power up to
four is usually sufficient. Thus, we follow this guideline to conduct
the RESET test. For the Breusch-Pagan test, the choice of predictors in
the auxiliary regression is left to the user
\citep{breusch_simple_1979}. But as \citet{waldman1983note} suggested,
it is a good choice for the set of auxiliary predictors in the
Breusch-Pagan test be the same as the White test. Thus, we include both
\(\boldsymbol{x}\) and \(\boldsymbol{x}^2\) in the auxiliary regression.

@@

For both the non-linearity pattern and the heteroskedasticity pattern,
the power curves of appropriate conventional tests are constantly
greater than the power curves of visual tests and other inappropriate
conventional tests. At \(log(E) = 1\) for non-linearity model, where the
residual departures are unnoticeable from the corresponding residual
plot, the RESET test has above 25\% of chance of rejecting \(H_0\),
while only a few visual tests would result in a rejection. Similarly, at
\(log(E) = 3\) for heteroskedasticity model, the BP test has around 25\%
power but the non-constant variance visual pattern is almost unobserved
from the residual plot and the power of the visual test is close to the
significance level at this point.
